### 2013-08-26 doc/dissertation/document/chapters/conclusion/conclusion.tex
moving estimator variance into an appendix

\chapter{Conclusions and Analysis\ }
\label{ch:conclusion}

@@

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations\ }
\label{sec:spn_conclusion}
This work demonstrates the first application of MCSA to the neutron
transport problem. To do this, we used the $SP_N$ form of the
Boltzmann transport equation. In this form, the transport equation
takes on a diffusion-like form where the angular character of the flux
is accounted for in moment terms much like those found in the $P_N$
equations. To meet the goals of this work and drive the research and
development of MCSA for application to light water reactor problems, a
difficult nuclear fuel assembly criticality calculation was used in a
set of numerical experiments. These calculations were enabled by
incorporating MCSA as a solution scheme into the k-eigenvalue solver
in the Exnihilo production neutronics code base at Oak Ridge National
Laboratory.

@@

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration Solutions for Navier-Stokes Equations\ }
\label{sec:nonlinear_conclusions}
This work presents the FANM method, a new nonlinear solution scheme
based on an inexact Newton method. To meet the goals of this work the
FANM method was characterized by applying it to the Navier-Stokes
equations. This research was enabled by implementing the FANM method
within the nonlinear solver sequence leveraged by the Drekar production
multiphysics code base being developed at Sandia National
Laboratories. In both convection and driven flow regimes we solved
three difficult benchmark problems for the Navier-Stokes
equations. Using the work in preconditioning from the research on the
$SP_N$ equations, it was found that the same preconditioning strategy
could be used to achieve convergence of the linear models generated at
each FANM iteration by applying an algebraic multigrid method.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel Monte Carlo Synthetic Acceleration\ }
\label{sec:parallel_mc_conclusions}
A new parallel MCSA algorithm was developed for this work to meet the
goal of demonstrating the improved scalability on leadership-class
hardware. It was found the MCSA can indeed be effectively parallelized
using the multiple-set overlapping-domain decomposition algorithm
borrowed from the reactor physics community. Using a neutron diffusion
problem, the parallel algorithm was verified to produce the same
results as two production Krylov methods.

>>>>

\chapter{Conclusion\ }
\label{ch:conclusion}

@@

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration Methods for the $SP_N$ Equations\ }
\label{sec:spn_conclusion}
This work demonstrates the first application of MCSA to the neutron
transport problem. To do this, we used the $SP_N$ form of the
Boltzmann transport equation. In this form, the transport equation
takes on a diffusion-like form where the angular character of the flux
is accounted for in moment terms much like those found in the $P_N$
equations. To meet the goals of this work and drive the research and
development of MCSA for application to light water reactor problems, a
difficult nuclear fuel assembly criticality calculation was used in a
set of numerical experiments. These calculations were enabled by
incorporating MCSA as a solution scheme into the k-eigenvalue solver
in the Exnihilo production neutronics code base at Oak Ridge National
Laboratory.

@@

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration Methods for the Navier-Stokes Equations\ }
\label{sec:nonlinear_conclusions}
This work presents the FANM method, a new nonlinear solution scheme
based on an inexact Newton method. To meet the goals of this work the
FANM method was characterized by applying it to the Navier-Stokes
equations. This research was enabled by implementing the FANM method
within the nonlinear solver sequence leveraged by the Drekar
production multiphysics code base being developed at Sandia National
Laboratories. In both convection and driven flow regimes we solved
three difficult benchmark problems for the Navier-Stokes
equations. Using the work in preconditioning from the research on the
$SP_N$ equations, it was found that the same preconditioning strategy
could be used to achieve convergence of the linear models generated at
each FANM iteration by applying an algebraic multigrid method.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel Monte Carlo Synthetic Acceleration Methods\ }
\label{sec:parallel_mc_conclusions}
A new parallel MCSA algorithm was developed for this work to meet the
goal of demonstrating the improved scalability on leadership-class
hardware. It was found the MCSA can indeed be effectively parallelized
using the multiple-set overlapping-domain decomposition algorithm
borrowed from the reactor physics community. Using a neutron diffusion
problem, the parallel algorithm was verified to produce the same
results as two production Krylov methods.

### 2013-08-26 doc/dissertation/document/chapters/introduction/introduction.tex
moving estimator variance into an appendix

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves beyond petascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to leverage this new hardware
\cite{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at this
new hardware.

Considering these physics and hardware issues, this goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems and a new method leveraging the
synthetic acceleration techniques to advance solution algorithms for
nonlinear problems in fluid flow. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. Once the linear and nonlinear methods have
been developed, we look forward to parallelizing them with the goal
that they may be leveraged competitively on leadership class computing
platforms.

To achieve these goals, three complementary research and development
activities are performed. First, Monte Carlo Sythetic Acceleration
(MCSA) is applied to the neutron transport problem in a light water
reactor calculation to develop the necessary preconditioning
techniques and understanding required to solve difficult
problems. Second, using these new techniques, a series of benchmarks
for the Navier-Stokes equations are solved in different flow regimes
to research the performance of the Monte Carlo algorithm for nonlinear
problems using the new Forward-Automated Newton-MCSA (FANM) algorithm
developed by this work. Third, a novel parallel algorithm based on
domain decomposed Monte Carlo methods for particle transport is
developed to fully parallelize the solution techniques developed to
solve the neutron transport and fluid flow problems on leadership
class hardware with scaling studies performed in this work using a
neutron diffusion problem.

@@

\subsection{Solutions for the $SP_N$ Equations}
\label{subsec:spn_motiviation}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, contain large numbers of degrees of freedom, or both,
requiring tremendous computational resources to generate an adequate
solution. Modern high fidelity deterministic methods for large scale
problems are commonly variants on the discrete ordinates ($S_N$)
method \cite{evans_denovo:_2010}. For fission reactor neutronics
simulations, the $S_N$ method requires potentially trillions of
unknown angular flux moments to be computed to achieve good accuracy
for the responses of interest
\cite{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Brantley and Larsen \cite{brantley_simplified_2000}, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
or $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment in many forms with both
variational and asymptotic derivations performed as a means of
verification \cite{olbrant_asymptotic_2013}. In addition, these
equations have been more rigorously studied as solution methods to MOX
fuel problems and have been shown to provide accurate solutions
\cite{brantley_simplified_2000}. With this mathematical literature to
provide a solid numerical footing for the method, we look at its
application to today's challenge problems in neutron transport for
fission reactor analysis. The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required.

@@

This leads us to then research the applicability of Monte Carlo
Synthetic Acceleration methods as a possible solution method for the
$SP_N$ equations. In addition, solving the $SP_N$ equations in this
way also breaks away from the $S_N$ forms of parallelism where spatial
parallelism is achieved by an efficient parallel sweep, angular
efficiency achieved by pipe-lining, and energy parallelism achieved by
decoupling the groups. With the $SP_N$ equations as a full matrix
system, we now can parallelize the problem as prescribed by the Monte
Carlo algorithm, which may be significantly more scalable than current
$S_N$ transport practices. In this work we therefore apply Monte Carlo
Synthetic Acceleration to the $SP_N$ equations and research its
applicability, performance, and convergence properties in the context
of a light water reactor system.

@@

Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
implicit nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods.

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over subspace methods due to
the fact that the physics operator must be explicitly formed. For many
applications including higher fidelity neutron transport
discretizations beyond the $SP_N$ equations discussed in the previous
section, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a discrete Monte Carlo solution
method is best suited for situations in which the operator is readily,
if not naturally, formed. Modern nonlinear methods meet this
requirement with Newton methods used in conjunction with Krylov
methods operating on a fully formed matrix for a robust solution
strategy. Furthermore, modern techniques exist that permit the
automatic construction of the Jacobian operator generated within a
Newton method based on the nonlinear residual evaluations, giving all
of the components necessary for a Monte Carlo solver to provide
value. We therefore devise a new nonlinear method in this work based
on the Monte Carlo Synthetic Acceleration algorithm and Newton's
method and research its properties, applicability, and performance in
the context of fluid flow problems.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems can be tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis for the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully implicit methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow for reactor analysis, then we are
motivated to advance the solution of complex physics problems
exploiting leadership class levels of parallelism.

For Monte Carlo Synthetic Acceleration methods to be viable at the
production scale for nuclear engineering applications, scalable
parallel implementations of the algorithm are required. Reviewing the
literature, Monte Carlo Synthetic Acceleration has yet to be
parallelized and the Neumann-Ulam Monte Carlo method on which it is
built has only been parallelized through history-level parallelism
with full domain replication \cite{alexandrov_efficient_1998}. In
order to solve large linear physics systems with Monte Carlo Synthetic
Acceleration, a domain decomposed parallel strategy is required. In
the formulation of a parallel Monte Carlo Synthetic Acceleration
algorithm, we recognize that the algorithm occurs in two stages, an
initial step performing a fixed point iteration and a secondary step
leveraging a Monte Carlo solver that is providing the
acceleration. The parallel aspects of both these components must be
considered. Therefore, in this work we will develop and implement
parallel algorithms for the fundamental random walk sequence and Monte
Carlo Synthetic Acceleration methods leveraging both the knowledge
gained from the general parallel implementations of subspace methods
reviewed in Appendix~\ref{ch:linear_problem} and modern parallel
strategies for domain decomposed Monte Carlo as developed by the
reactor physics community.

@@

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With linear and
nonlinear solution schemes based on Monte Carlo techniques, such
issues are potentially alleviated by statistical arguments. In the
case of soft failures, isolated floating point errors in Monte Carlo
simulation are absorbed within tally statistics while completely
losing hardware during a hard failure is manifested as a high variance
event where some portion of the Monte Carlo histories are lost. These
stochastic methods are a paradigm shift from current deterministic
solver techniques that will in their current form potentially suffer
greatly from the non-deterministic behavior expected from the next
generation of machines.

In addition to resiliency concerns, the memory restrictions on future
hardware will hinder modern solvers that derive their robustness from
using large amounts of memory. Stochastic methods that are formulated
to use less memory than conventional methods will serve to alleviate
some of this pressure. New parallel strategies that may
be implemented with stochastic methods could offer a new avenue for
leveraging the expected levels of high concurrency in machines
operating at hundreds of petaflops and beyond.

%%---------------------------------------------------------------------------%%
\section{Research and Development Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\cite{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \cite{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\cite{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
\cite{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the physics
operator need not be symmetric or positive-definite, (2) the
stochastic nature of the solution method provides a natural solution
to some aspects of the issue of resiliency, and (3) is amenable to
parallelization using modern methods developed by the transport
community \cite{wagner_hybrid_2010}. The development of MCSA as a
general solution technique and the development of a parallel MCSA
method are new and unique features of this work, providing a framework
with which other issues such as resiliency or performance may be
addressed in the future.

In addition to advancements in the parallel solution of linear
problems, nonlinear problems may also benefit from a general and
parallel MCSA scheme. In the nuclear engineering community, nonlinear
problems are often addressed by either linearizing the problem or
building a segregated scheme and using traditionally iterative or
direct methods to solve the resulting system
\cite{pletcher_computational_1997}. In the mathematics community,
various Newton methods have been popular
\cite{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
(JFNK) schemes \cite{knoll_jacobian-free_2004} have been utilized in
multiple physics architectures and advanced single physics codes
\cite{gaston_parallel_2009}. The benefits of JFNK schemes are that the
Jacobian is never formed, simplifying the implementation, and a Krylov
solver is leveraged (typically GMRES or a method based on Conjugate
Gradient), providing excellent convergence properties for
well-conditioned and well-scaled systems. However, there are two
potential drawbacks to these methods for high fidelity predictive
simulations: (1) the Jacobian is approximated by a first-order
differencing method on the order of machine precision such that this
error can grow beyond that of those in a fine-grained system
\cite{kelley_iterative_1995} and (2) for systems that are not
symmetric positive-definite (which will be the case for most
multiphysics systems and certainly for most preconditioned systems)
the Krylov subspace generated by the GMRES solver may become
prohibitively large \cite{knoll_newton-krylov_1995}.

In this work, the Forward-Automated Newton-MCSA (FANM) method is
developed as new nonlinear solution method. The key features of FANM
are: full Jacobian generation using modern Forward Automated
Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
as the inner linear solver. This method has several attractive
properties. First, the first-order approximation to the Jacobian used
in JFNK type methods is eliminated by generating the Jacobian
explicitly with the model equations through FAD. Second, the Jacobian
need not be explicitly formed by the user but is instead automated
through FAD; this eliminates the complexity of hand-coding derivatives
and has also been demonstrated to be more efficient computationally
than evaluating difference derivatives. Third, unlike GMRES, MCSA does
not build a subspace during iterations. Although the Jacobian must be
explicitly formed to use MCSA, for problems that take more than a few
GMRES iterations to converge the size of the Krylov subspace will
typically grow beyond that of the Jacobian. Finally, using MCSA to
solve the linear model provides its benefits of potential parallelism
and iterative performance.

@@

The goal of this work is to improve the iterative performance and
parallel scalability of solutions to discrete linear and nonlinear
transport problems by researching and developing a new set of domain
decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
iterative performance and parallel scalability will be demonstrated in
a wide variety of numerical experiments using both simple model
neutron transport problems and neutron transport and fluid flow
problems generated by production physics codes in both simple and
complex geometries.

>>>>

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves beyond petascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to utilize this new hardware
\cite{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at the
next generation of machines.

Considering these physics and hardware issues, this goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems and a new method leveraging the
synthetic acceleration techniques to advance solution algorithms for
nonlinear problems in fluid flow. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. Once the linear and nonlinear methods have
been developed, we look forward to parallelizing them with the goal
that they may be leveraged competitively on leadership-class computing
platforms.

To achieve these goals, three complementary research and development
activities are performed. First, Monte Carlo Sythetic Acceleration
(MCSA) is applied to the neutron transport problem in a light water
reactor calculation to develop the necessary preconditioning
techniques and understanding required to solve difficult
problems. Second, using these new techniques, a series of benchmarks
for the Navier-Stokes equations are solved in different flow regimes
to research the performance of the Monte Carlo algorithm for nonlinear
problems using the new Forward-Automated Newton-MCSA (FANM) algorithm
developed by this work. Third, a novel parallel algorithm based on
domain decomposed Monte Carlo methods for particle transport is
developed to fully parallelize the new solution techniques on
leadership class hardware with scaling studies performed in this work
using a neutron diffusion problem.

@@

\subsection{Solutions for the $SP_N$ Equations}
\label{subsec:spn_motiviation}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, contain large numbers of degrees of freedom, or both,
requiring tremendous computational resources to generate an adequate
solution. Modern high fidelity deterministic methods for large scale
problems are often variants on the discrete ordinates ($S_N$) method
\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
the $S_N$ method requires potentially trillions of unknown angular
flux moments to be computed to achieve good accuracy for the responses
of interest \cite{slaybaugh_acceleration_2011}. Other forms of the
transport problem, including the $P_N$ method, take on a simpler form
than the more common $S_N$ methods, but lack in accuracy when compared
while still requiring considerable computational resources for
solutions in multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc, multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Brantley and Larsen \cite{brantley_simplified_2000}, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
or $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification
\cite{olbrant_asymptotic_2013}. In addition, these equations have been
more rigorously studied as solution methods to MOX fuel problems and
have been shown to provide accurate solutions
\cite{brantley_simplified_2000}. With this mathematical literature to
provide a solid numerical footing for the method, we look at its
application to today's challenge problems in neutron transport for
fission reactor analysis. The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required.

@@

This leads us to then research the applicability of Monte Carlo
Synthetic Acceleration methods as a possible solution method for the
$SP_N$ equations. In addition, solving the $SP_N$ equations in this
way also breaks away from the $S_N$ forms of parallelism where spatial
parallelism is achieved by an efficient parallel sweep, angular
efficiency achieved by pipe-lining, and energy parallelism achieved by
decoupling the groups. With the $SP_N$ equations as a full matrix
system, we now can parallelize the problem as prescribed by the Monte
Carlo algorithm, which may be significantly more scalable than current
$S_N$ transport practices. In this work we apply Monte Carlo Synthetic
Acceleration to the $SP_N$ equations and research its applicability,
performance, and convergence properties in the context of a light
water reactor system.

@@

Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
implicit nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods \cite{gaston_parallel_2009}.

In the context of solving standalone linear problems, Monte Carlo
methods do not provide significant merit over subspace methods due to
the fact that the physics operator must be explicitly formed. For many
applications including higher fidelity neutron transport
discretizations beyond the $SP_N$ equations discussed in the previous
section, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a discrete Monte Carlo solution
method is best suited for situations in which the operator is readily,
if not naturally, formed. Modern nonlinear methods meet this
requirement with Newton methods used in conjunction with Krylov
methods operating on a fully formed matrix for a robust solution
strategy. Furthermore, modern techniques exist that permit the
automatic construction of the Jacobian operator generated within a
Newton method based on the nonlinear residual evaluations, giving all
of the components necessary for a Monte Carlo solver to provide
value. In this work we devise a new nonlinear method in this work
based on the Monte Carlo Synthetic Acceleration algorithm and Newton's
method and research its properties, applicability, and performance in
the context of fluid flow problems.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems can be tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis for the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub
\cite{u.s._department_of_energy_casl_2011}. CASL used the Drekar
multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully implicit methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow for reactor analysis, then we are
motivated to advance the solution of complex physics problems
exploiting leadership class levels of parallelism.

For Monte Carlo Synthetic Acceleration methods to be viable at the
production scale for nuclear engineering applications, scalable
parallel implementations of the algorithm are required. Reviewing the
literature, Monte Carlo Synthetic Acceleration has yet to be
parallelized and the underlying Neumann-Ulam Monte Carlo method has
only been parallelized through history-level parallelism with full
domain replication \cite{alexandrov_efficient_1998}. In order to solve
large linear physics systems with Monte Carlo Synthetic Acceleration,
a domain decomposed parallel strategy is required. In the formulation
of a parallel Monte Carlo Synthetic Acceleration algorithm, we
recognize that the algorithm occurs in two stages, an initial step
performing a fixed point iteration and a secondary step leveraging a
Monte Carlo solver that is providing the acceleration. The parallel
aspects of both these components must be considered. Therefore, in
this work we will develop and implement parallel algorithms for the
fundamental random walk sequence and Monte Carlo Synthetic
Acceleration methods leveraging both the knowledge gained from the
general parallel implementations of subspace methods reviewed in
Appendix~\ref{ch:linear_problem} and modern parallel strategies for
domain decomposed Monte Carlo as developed by the reactor physics
community.

@@

The end result of these hardware changes is that the larger number of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With linear and
nonlinear solution schemes based on Monte Carlo techniques, such
issues are potentially alleviated by statistical arguments. In the
case of soft failures, isolated floating point errors in the Monte
Carlo simulation are absorbed within tally statistics. Completely
losing memory during a hard failure is treated as a high variance
event where some portion of the Monte Carlo histories and subsequently
the solution are lost with other histories maintained by replicating
the problem and combining the results through superposition. These
stochastic methods are a paradigm shift from modern deterministic
solver techniques that will, in their current form, potentially suffer
greatly from the non-deterministic behavior expected from the next
generation of machines.

%%---------------------------------------------------------------------------%%
\section{Research and Development Outline}
\label{sec:research_outline}
For some time, the particle transport community has utilized Monte
Carlo methods for the solution of transport problems
\cite{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \cite{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\cite{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
\cite{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the physics
operator need not be symmetric or positive-definite, (2) the
stochastic nature of the solution method provides a natural solution
to some aspects of the issue of resiliency, and (3) is amenable to
parallelization using modern methods developed by the transport
community \cite{wagner_hybrid_2010}. The development of MCSA as a
general solution technique and the development of a parallel MCSA
method are new and unique features of this work, providing a framework
with which other issues such as resiliency or performance may be
addressed in the future.

In addition to advancements in the parallel solution of linear
problems, nonlinear problems may also benefit from a general and
parallel MCSA scheme. In the nuclear engineering community, nonlinear
problems are often addressed by either linearizing the problem or
building a segregated scheme and using traditionally iterative or
direct methods to solve the resulting system
\cite{pletcher_computational_1997}. In the mathematics community,
various Newton methods have been popular
\cite{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
(JFNK) schemes \cite{knoll_jacobian-free_2004} have been utilized in
multiple physics architectures and advanced single physics codes
\cite{gaston_parallel_2009}. The benefits of JFNK schemes are that the
Jacobian is never formed, simplifying the implementation, and a Krylov
solver is leveraged (typically GMRES or a method based on Conjugate
Gradient), providing excellent convergence properties for
well-conditioned and well-scaled systems. However, there are two
potential drawbacks to these methods for high fidelity predictive
simulations: (1) the Jacobian is approximated by a first-order
differencing method on the order of machine precision such that this
error can grow beyond that of those in a poorly-scaled system
\cite{kelley_iterative_1995} and (2) for systems that are not
symmetric positive-definite (which will be the case for most
multiphysics systems and certainly for most preconditioned systems)
the Krylov subspace generated by the GMRES solver may become
prohibitively large \cite{knoll_newton-krylov_1995}.

In this work, the Forward-Automated Newton-MCSA (FANM) method is
developed as new nonlinear solution method. The key features of FANM
are: full Jacobian generation using modern Forward Automated
Differentiation (FAD) methods, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated with the Jacobian fully formed. Second, the Jacobian need
not be explicitly formed by the user but is instead automated through
FAD; this eliminates the complexity of hand-coding derivatives and has
also been demonstrated to be more efficient computationally than
evaluating difference derivatives
\cite{bartlett_automatic_2006}. Third, unlike GMRES, MCSA does not
build a subspace during iterations. Although the Jacobian must be
explicitly formed to use MCSA, for problems that take more than a few
GMRES iterations to converge the size of the Krylov subspace will
typically grow beyond that of the Jacobian. Finally, using MCSA to
solve the linear model provides its benefits of potential parallelism
and iterative performance.

@@

The goal of this work is to improve the iterative performance and
parallel scalability of solutions to discrete linear and nonlinear
transport problems by researching and developing a new set of domain
decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
iterative performance and parallel scalability will be demonstrated in
a wide variety of numerical experiments using both simple model
transport problems and neutron transport and fluid flow problems
generated by production physics codes in both simple and complex
geometries.
